Consistency and variability in the development of children's drawings of object categories

### 2021-06-26 writing/Manuscript_June2021.Rmd
smoothing edits

abstract: |
 Childen's drawings of common object categories become dramatically more recognziable across development. What are the major factors that explain this developmental change? Here, we examined the degree to which these developmental changes in recognizibility vary across different drawing tasks (i.e. drawing from observation vs. from memory), geographical locations (San Jose, US vs. Beijing, China),and with children's tracing abilities. To do so, we collecting digital drawings of object categories (e.g., cat, airplane) from 4-9 year-olds (N=253). Overall, we show broad consistency in these developmental trajectories acros both drawing tasks and these two geographical locations, and we find that children's tracings abilites are good predictors of the recognizability of the drawings that they produce. In addition, we find that children recruited near Beijing, China producedmore recognizable drawings but showed similar tracing abilities to childen recruited near San Jose, CA. Overall, this work suggests that the developmental trajectory of children's drawings are remarkably consistent and not easily explainable by changes in their visuomotor control.

@@

## Measuring Tracing Accuracy
As in [@long2018drawings], we used an automated procedure for evaluating how accurately participants performed the tracing task, validated against empirical judgments of tracing quality. We decomposed tracing accuracy into two components: a shape error component and a spatial error component. Shape error reflects how closely the participant’s tracing matched the contours of the target shape; the spatial error reflects how closely the location, size, and orientation of the participant’s tracing matched the target shape.

@@

We used the same tracing quality ratings to obtained in [@long2018drawings] to assign weights to each of their error terms; adult observers ($N$=70) rated 1325 tracings (i.e., 50-80 tracings per shape per age) and evaluated “how well the tracing matches the target shape and is aligned to the position of the target shape” on a 5-point scale. An ordinal regression mixed-effects model to predict these 5-point ratings, which contained correlation distance, translation, rotation, scaling, and shape identity (square vs. star) as predictors, with random intercepts for rater.  This model yielded parameter estimates that could then be used to score each tracing in the dataset; we averaged scores for both shapes within session to yield a single tracing score for each participant.

@@

### Human recognition scores
We assessed the recognizability of each drawing via an online recognition experiment. Adult participants based in the U.S. were recruited via Prolific for a 15 minute experiment,compensated at $14/hour, and recognized a random subset of around 140 drawings that were balanced with respect to age, category, and site. Each drawing was recognized by 10 participants. Participants were shown a were shown these drawings in a random order and asked "What does this look like?". Participants selected an answer from the twelve categories that children were prompted to draw and were were encouraged to guess. No participants were excluded from analysis for missing a catch trial (a free response describing the task that were completing). These binary recognition scores were then averaged across participants to yield a recognition score (e.g., recognized by 80% of participants) for each drawing.

@@

To assess our main hypotheses, we fit linear mixed effects models to the human recogniton scores to assess the factors that influenced the recognizability of the drawings that children produced. A first linear mixed effect model was fit to the recognizability scores assigned to each drawing, including fixed effects of children's age (in years), geographical site (San Jose vs. Beijing), and drawing task (drawing from perception vs. memory) and the three way interaction between these key variables. We initially planned to include random slopes for the effect of drawing task on each child (as this varied within-subjects), and random slopes for the effect the full three-way interactions between task, age, and site on each category. However, all models with this complicated random effects structure failed to converge, and the reported models the maximal random effects structure that did converge.

In a secondary analysis, we aimed to understand the degree to which any of the above effects were mediated by (1) children's tracing abilities and (2) the amount of effort that children expended while drawings. We thus ran the same main model while also now including fixed effects of children's estimated tracing score (see above), the time children spent drawing (in seconds), the mean intensity of the drawing (i.e. percentage of non-white pixels), and the number of strokes children used. All predictors were scaled to have a mean of 0 and a standard deviation of 1. Finally, we also assessed the degree to which tracing ability development differed by geographical site, where tracing scores were modeled as a function of age (in years), site, and their interaction, with random intercepts for each participant.

This analysis plan was pre-registered at https://osf.io/qymjr/; however, we had initially planned to use automated recognition scores as our main dependent variable, following prior work [@long2021parallel]; however, we found that these automated recognition scores were only modestly correlated with human recognition scores ($r$ = `r round(human_v_model$estimate,2)`, $t$ = `r round(human_v_model$statistic,2)`, p < .001); furthermore, descriptive plots of the automated recognition scores revealed a difference between the two drawing tasks that was not evident in the human data. To be conservative, we thus fit all of our models to the human recognition data but have made all model classifications available for future researcs.

@@

# Results
Overall, we found relative consistency in the developmental trajectory of children's drawings. Figure \ref{fig:main-results} shows the recognizability of children's drawings at each age as a function of the drawing task and the geographical site that they were located in; All model coefficients can be found in Table 1.  First, we found consistent and steady changes in the recognizability of children's drawings across this age range, replicating prior work with observational datasets collected at a free-standing kiosk [@long2021parallel]. Critically, these changes in recognizability did not depend on the drawing task children completed: children's drawings were equally recognizable when they were asked to "Draw this rabbit as it looks in the picture" and they were simply asked "What about a rabbit? Can you draw a rabbit." While we anticipated that older children might produce more recognizable drawings when drawing from observation -- as do adults when presented with canonical photographs of object categories [@yang2021visual] -- this was not the case.

However, we did observe a main effect of geographical site: Children who were recruited near Beijing, China tended to produce more recognizable drawings than children who were recruited near San Jose, USA. In a second set of analyses, we aimed to understand the nature of this difference in recognizability. We hypothesized that any differences in the recognizability of children's drawings across sites would be mediated by the amount of effort children expended while drawing or their tracing abilities.  We thus first examined how effort and tracing abilities varied across task, age, and geographical location. Figure \ref{fig:effort-covariates} shows three effort covariates - average intensity, number of strokes used, and time spent drawing -- measured for each drawing as a function of children's age, drawing task, and geographical site: effort covariates measured for each drawing; children tended to spend more time drawing when asked to draw "the [rabbit] as it looks in the picture", and children recruited outside Beijing tended to spend more time drawing overall. In contrast, children's tracing abilities did not vary across geographical site: Figure \ref{fig:tracing} shows the average tracing score at each age at each site. While children's tracing abilities increased steadily with age, as in prior work [@long2021parallel], children recruited at both sites produced equally good shape tracings.

Next, we included children's average tracing scores and effort covariates measured for each drawing (average intensity, number of strokes used, and time spent drawing) as fixed effects into a second linear mixed-effects model with the same fixed and random effects structure as used above. If these effort covariates accounted the differences between sites that we observed, we reasoned that we should no longer observe a main effect of geographical location on drawing recognizability. Contra this hypothesis, however, we still observed a significant effect of geographical site (see Table 2), despite the fact that individual children's tracing abilities were strong predictors of how recognizable their drawings were. In addition, the amount of time a child spent drawing  was a significant negative predictor of recognizability -- indicating that the amount of effort that children expended did not necessarily result in a drawing that was more recognizable to others.

Thus, these results suggest that these two groups of children differ in their ability to produce recognizable drawings of object categories in a way that is not esaily explainable by effort covariates or tracing abilities, contra previous accounts. In a third set of analyses, we examined how the developmental trajectory of the recognizability of children's drawings may differ for different object categories. Intuitively, some object categories (e.g., cat) may be easier to draw than others (e.g., watch), resulting in shallower or steeper changes in recognizability over age. However, we had no strong hypotheses about how these trajectories might additionally vary with geographical locations.

Figure \ref{fig:item-effects} shows the recognizability of children's drawings for each of the 12 categories a function of geographical location and children's age and reveal considerable variability.  For example, certain categories (e.g., cats and rabbits) were overall more difficult for children to depict such that they were distinct from one another, particularly for younger children. In addition,  children recruited near Beijing produced more recognizable drawings of certain categories at all ages -- including airplanes, birds, and rabbits -- while older children recruited near San Jose produced more recognizable drawings of bikes.  Figure \ref{fig:example-drawings} shows recognizable, randomly sampled example drawings from 6-year-olds at each category, site, and condition.

@@

Instead, we found that the recognizability of children's drawings differed between our two different geographical sites -- Beijing, China and San Jose, USA  Why might this be the case?  Our initial hypothesis was that children who had daily practice with a non-orthographic, complex writing system would exhibit greater dexterity in our tracing assessment task -- and that this dexterity could be a factor in how recogizable their drawings were.  However, children across the two sites showed equivalent tracing abilities -- despite the fact that children's individual tracing scores were strong predictors of the recognizability of the drawings that they produced. Thus, this finding both confirms the idea that children's visuomotor abilities strongly influence the drawings they produce yet further suggests that it is far from the only factor that does so.

>>>>

abstract: |
 Childen's drawings of common object categories become dramatically more recognziable across development. What are the major factors that explain this developmental change? Here, we examined the degree to which these developmental changes in recognizibility vary across different drawing tasks (i.e. drawing from observation vs. from memory), geographical locations (San Jose, US vs. Beijing, China), and with children's tracing abilities. To do so, we collecting digital drawings of object categories (e.g., cat, airplane) from 4-9 year-olds (N=253). Overall, we show broad consistency in these developmental trajectories acros both drawing tasks and these two geographical locations, and we find that children's tracings abilites are good predictors of the recognizability of the drawings that they produce. In addition, we find that children recruited near Beijing, China produced more recognizable drawings but showed similar tracing abilities to childen recruited near San Jose, USA. Overall, this work suggests that the developmental trajectory of children's drawings are remarkably consistent and not easily explainable by changes in their visuomotor control or working memory capacity.

@@

## Measuring Tracing Accuracy
As in @long2021parallel, we used an automated procedure for evaluating how accurately participants performed the tracing task, validated against empirical judgments of tracing quality. We decomposed tracing accuracy into two components: a shape error component and a spatial error component. Shape error reflects how closely the participant’s tracing matched the contours of the target shape; the spatial error reflects how closely the location, size, and orientation of the participant’s tracing matched the target shape.

@@

We used the tracing quality ratings to obtained in @long2021parallel to assign weights to each of their error terms; adult observers ($N$=70) rated 1325 tracings (i.e., 50-80 tracings per shape per age) and evaluated “how well the tracing matches the target shape and is aligned to the position of the target shape” on a 5-point scale. An ordinal regression mixed-effects model to predict these 5-point ratings, which contained correlation distance, translation, rotation, scaling, and shape identity (square vs. star) as predictors, with random intercepts for rater. This model yielded parameter estimates that could then be used to score each tracing in the dataset; we averaged scores for both shapes to yield a single tracing score for each participant.

@@

### Human recognition scores
We assessed the recognizability of each drawing via an online recognition experiment. Adult participants based in the U.S. were recruited via Prolific for a 15 minute experiment, compensated at $14/hour, and asked to recognize a random subset of around 140 drawings that were balanced with respect to age, category, and site. Each drawing was recognized by 10 participants. Participants were shown these drawings in a random order and asked "What does this look like?". Participants selected an answer from the twelve categories that children were prompted to draw and were encouraged to guess. No participants were excluded from analysis for missing a catch trial (a free response describing the task that were completing). These binary recognition scores were then averaged across participants to yield a recognition score for each drawing.
(e.g., recognized by 80% of participants).

@@

To assess our main hypotheses, we fit linear mixed effects models to the human recognition scores to assess the factors that influenced the recognizability of the drawings that children produced. A first linear mixed effect model was fit to the recognizability scores for each drawing, including fixed effects of children's age (in years), geographical site (San Jose vs. Beijing), and drawing task (drawing from perception vs. memory) and the three way interaction between these key variables. We initially planned to include random slopes for the effect of drawing task on each child (as this varied within-subjects), and random slopes for the effect the full three-way interactions between task, age, and site on each category. However, all models with this complicated random effects structure failed to converge, and the reported models use the maximal random effects structure that did converge (see corresponding codebase).

In a secondary analysis, we aimed to understand the degree to which any of the above effects were mediated by (1) children's tracing abilities and (2) the amount of effort that children expended while drawings. We thus ran the same main model while also now including fixed effects of children's estimated tracing score (see Measuring Tracing Abilities), the time children spent drawing (in seconds), the mean intensity of the drawing (i.e. percentage of non-white pixels), and the number of strokes children used. All predictors were scaled to have a mean of 0 and a standard deviation of 1. Finally, we also assessed the degree to which tracing ability development differed by geographical site, where tracing scores were modeled as a function of age (in years), site, and their interaction, with random intercepts for each participant.

This analysis plan was pre-registered at https://osf.io/qymjr. While we had initially planned to use automated recognition scores as our main dependent variable, following prior work [@long2021parallel],  we found that these automated recognition scores were only modestly correlated with human recognition scores ($r$ = `r round(human_v_model$estimate,2)`, $t$ = `r round(human_v_model$statistic,2)`, p < .001); furthermore, descriptive plots of the automated recognition scores revealed a overall difference between the two drawing tasks that was not evident in the human recognition data. To be conservative, we thus fit all of our models to the human recognition data but have made all model classifications available for future research.

@@

# Results
Overall, we found relative consistency in the developmental trajectory of children's drawings. Figure \ref{fig:main-results} shows the recognizability of children's drawings at each age as a function of the drawing task and the geographical site that they were located in; All model coefficients can be found in Table 1.  First, we found steady changes in the recognizability of children's drawings across this age range, replicating prior work with an observational dataset collected at a free-standing kiosk [@long2021parallel]. Critically, these changes in recognizability did not depend on the drawing task children completed: children's drawings were equally recognizable when they were asked to "Draw this rabbit as it looks in the picture" and they were simply asked "What about a rabbit? Can you draw a rabbit." While we anticipated that older children might produce more recognizable drawings when drawing from observation -- as do adults when presented with canonical photographs of object categories [@yang2021visual] -- this was not the case.

However, we did observe a main effect of geographical site: Children who were recruited near Beijing, China tended to produce more recognizable drawings than children who were recruited near San Jose, USA. In a second set of analyses, we aimed to understand the nature of this difference in recognizability. We hypothesized that any differences in the recognizability of children's drawings across sites would be mediated by the amount of effort children expended while drawing or their tracing abilities.  We thus first examined how effort and tracing abilities varied across task, age, and geographical location. Figure \ref{fig:effort-covariates} shows three effort covariates - average intensity, number of strokes used, and time spent drawing -- measured for each drawing as a function of children's age, drawing task, and geographical site.  Cchildren tended to spend more time drawing when asked to draw "the [rabbit] as it looks in the picture", and children recruited outside Beijing tended to spend more time drawing overall. In contrast, children's tracing abilities did not vary across geographical site: Figure \ref{fig:tracing} shows the average tracing score at each age at each site. While children's tracing abilities increased steadily with age, as in prior work [@long2021parallel], children recruited at both sites produced equally good tracings.

We then included children's average tracing scores and effort covariates measured for each drawing (average intensity, number of strokes used, and time spent drawing) as fixed effects into a second linear mixed-effects model (described in Statistical Models). If the amount of effort spent or children's tracing abilities accounted the differences between sites that we observed, we reasoned that we should no longer observe a main effect of geographical location on drawing recognizability. Contra this hypothesis, however, we still observed a significant effect of geographical site (see Table 2), despite the fact that individual children's tracing abilities were strong predictors of how recognizable their drawings were. In addition, the amount of time a child spent drawing was in fact a significant negative predictor of recognizability -- indicating that the amount of effort that children expended did not necessarily result in a drawing that was more recognizable to others.

Thus, these results suggest that these two groups of children differ in their ability to produce recognizable drawings of object categories in a way that is not easily explainable by effort covariates or tracing abilities. In a third set of exploratory analyses, we examined how the developmental trajectory of the recognizability of children's drawings differ for different object categories. Intuitively, some object categories (e.g., cat) may be easier to draw than others (e.g., watch), resulting in shallower or steeper changes in recognizability over age. However, we had no strong hypotheses about how these trajectories might additionally vary with geographical locations.

Figure \ref{fig:item-effects} shows the recognizability of children's drawings for each of the 12 categories a function of geographical location and children's age and reveal considerable variability.  For example, certain categories (e.g., cats and rabbits) were overall more difficult for children to depict such that they were distinct from one another, particularly for younger children. In addition, children recruited near Beijing produced more recognizable drawings of certain categories at all ages -- including airplanes, birds, and rabbits -- while older children recruited near San Jose produced more recognizable drawings of bikes.  Figure \ref{fig:example-drawings} shows recognizable, randomly sampled example drawings from 6-year-olds at each category, site, and condition. These exploratory analyses thus reveal systematic differences between object categories and geographical site that do not seem easily attributable to differences in effort.

@@

Instead, we found that the recognizability of children's drawings differed between our two different geographical sites -- Beijing, China and San Jose, USA  Why might this be the case?  Our initial hypothesis was that children who had daily practice with a non-orthographic, complex writing system would exhibit greater dexterity in our tracing assessment task -- and that this dexterity could be a factor in how recogizable their drawings were.  However, children across the two sites showed equivalent shape tracing abilities -- despite the fact that children's individual tracing scores were strong predictors of the recognizability of the drawings that they produced. Thus, this finding both confirms the idea that children's visuomotor abilities strongly influence the drawings they produce yet further suggests that it is far from the only factor that does so.
